---
title: "Question 4: Effects of Flight Phase"
format: pdf 
#include-in-header: custom-latex.sty
editor: visual
---

## ***Q4 - Flight Phase & Bird Strikes:***

### Do all flight phases have the same average number of birds struck per incident? If not, which flight phase is correlated with the highest average number of bird strikes? Is there a casual relationship between flight phase and bird strikes?

### 

## Analysis

To begin, I will utilize basic EDA approaches by plotting summary statistics about the data I plan to analyze. I'll look into the means of `NumberStruckActual` by flight phase using a bar chart, as well as explore the distribution of `NumberStruckActual` utilizing a violin plot. To answer the second part of this question, I plan to utilize a Bayesian Generalized Linear Model (GLM). Viewing these averages and understanding the distribution will allow me to set better priors for the GLM. I plan to utilize normal, weakly informative priors with a poisson family, as the outcome data is count data. I also plan to use a log link function to assure that the outcomes are positive, as it is not reasonable to have a negative count of birds struck. Before running the GLM, I will build a DAG to explore the causal relationships present in this dataset. Building this DAG will allow me to control for any confounders and give me the ability to make a causal estimate. Once the EDA is complete, I plan to build multiple GLM models with different families and priors to test the different effects on the model. Afterwards, I will utilize prior predictive checks and look at the ESS and posterior intervals and distributions to select the best model.

This analysis will allow me to fully answer all parts of this question. Using a Bayesian approach will allow me to utilize prior knowledge and deeply assess the model. I plan to use variables that I believe are correlated with bird strikes including flight information and geographical information.

```{r setup, include=FALSE, echo = FALSE}
# Load libraries
library(tidyverse)
library(ggdag)
library(rstanarm)
library(bayesplot)
library(fastDummies)
library(corrplot)
library(dagitty)

# Load data
data <- read.csv("Bird_strikes.csv", stringsAsFactors = TRUE)

# Replace NAs with 0
data[is.na(data)] <- 0


```

## EDA

```{r overview, include = TRUE, echo = FALSE}
# Summarize bird strikes by flight phase
bird_strikes_summary <- data %>%
  group_by(FlightPhase) %>%
  summarise(
    MeanStrikes = mean(NumberStruckActual, na.rm = TRUE),
    VarianceStrikes = var(NumberStruckActual, na.rm = TRUE),
    Count = n()
  )

# Bar plot of mean bird strikes by flight phase
ggplot(bird_strikes_summary, aes(x = FlightPhase, y = MeanStrikes, fill = FlightPhase)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Average Bird Strikes by Flight Phase",
    x = "Flight Phase",
    y = "Average Number of Bird Strikes") + 
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(data, aes(x = FlightPhase, y = NumberStruckActual, fill = FlightPhase)) +
  geom_violin() +
  labs(
    title = "Distribution of Bird Strikes by Flight Phase",
    x = "Flight Phase",
    y = "Number of Bird Strikes") + 
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))



```

## Part A. Do all flight phases have the same average number of birds struck per incident?

```{r, include = TRUE, echo = FALSE}
anova <- aov(NumberStruckActual ~ FlightPhase, data)
summary(anova)

```

The P-value from this ANOVA confirms that the different levels of `FlightPhase` have statistically significant differences at the p=0.05 level. Because of this, we will do a further exploration to determine which `FlightPhase` has the highest average number of bird strikes.

## Part B. Which flight phase is correlated with the highest average number of bird strikes? Is there a casual relationship between flight phase and bird strikes?

### DAG Creation & Quantification

In this DAG, we visualize and explore the relationships between the variables `NumberStruckActual`, `FlightPhase`, `Altitude`, `ConditionsSky`, `Effect`, `Damage`, and `Origin`. Ultimately, we are aiming to quantify the direct effect of `FlightPhase` on `NumberStruckActual`, making `FlightPhase` our expose variable and `NumberStruckActual` our outcome variable. There exists a chain from `FlightPhase` -\> `Altitude` -\> `NumberStruckActual`, making `Altitude` a mediator between `FlightPhase` and `NumberStruckActual`. Because we are interested in the *direct* effect, we must control for `Altitude` in our model. To do so, we will include `Altitude` as an additional predictor in our GLM.

```{r DAG Output, include = TRUE, echo = FALSE}
# Construct DAG

dag <- ggdag::dagify(
    NumberStruckActual ~ Altitude + FlightPhase + Origin,
    ConditionsSky ~ Origin, 
    Altitude ~ FlightPhase, 
    Damage ~ NumberStruckActual, 
    Effect ~ Damage, 
    exposure = "FlightPhase", 
    outcome = "NumberStruckActual", 
    labels = c(FlightPhase = "FlightPhase", 
           ConditionsSky = "ConditionsSky", 
           Effect = "Effect", 
           Damage = "Damage", 
           OriginState = "OriginState", 
           NumberStruckActual = "NumberStruckActual", 
           Altitude = "Altitude", 
           Origin = "Origin")
)


# Plot the DAG
ggdag(dag, node_size = 18, layout = "circle", stylized = TRUE, use_labels = "label", text = "False", 
      node = "True", label_size = 2.5) +
  theme_dag()
  labs(title = "DAG: Confounding Variables for Bird Strikes")


print(adjustmentSets(dag, effect = "direct"))
```

### Bayesian Parameter Estimation + Causal Inference

To complete Bayesian parameter estimation, we tested a dew different priors and families. All priors tested were weakly informative, as our prior knowledge of bird strikes is limited. The priors we tested included a normal prior with a $\mu$ of 0 and a $\sigma$ of 2.5, a student-t prior with 3 degrees of freedom, a location of 0, and a scale of 2.5. Using the student-t may allow us to better account for outliers, which we know are present in this dataset. For the intercept priors, we used both normal and student-t priors, although we allowed for higher variances/scales, as we wanted to allow for more variance in the intercepts. We selected distributions centered at 0, because we believe that it is reasonable for flight phases and altitude to have a positive or negative impact on `NumberStruckActual`, and that impact is most likely close to 0. From exploring our data, we know that most bird strikes involve very few birds, so we are not expecting parameters to be very large in magnitude. We used a log link function, as we want to ensure that all of our outcomes are positive. We tested two families, Poisson and negative binomial. equal to poisson, as this is count data. An assumption of a Poisson distribution is that the means and variances are the same. We know that there is a lot of variance in this data, due to several large outliers, so a Poisson distribution may not be the best fit. The alternative we attempted, negative binomial, is good for cases where the data is overdispersed, or the variance is much greater than the mean.

After running the models, the one I use is as follows: The priors that were chosen are weakly informative priors of a normal distribution, with a mean of 0 and standard deviation of 2.5. The family is a negative binomial regression model with a log link function, to account for overdispersion. The outcome variable is `NumberStruckActual`, and the predictors are `FlightPhase` and `Altitude`. The link function is a log function as we to ensure values are positive. The prior is centered at 0, as it is assumes that most predictors will have little or no influence on the outcome but allowing for a moderately large variation of 2.5. The prior intercept is a normal prior with a mean of 0 and variation of 5 so the model can have a larger baseline counts of bird strikes which shows more uncertainty. The model has 4 chains with 8000 iterations so that we can have a large posterior sample size and have more robust posterior estimates.

For the sake of the runtime it takes to render this PDF, only the one selected model is left uncommented.

```{r}
#install.packages("rstanarm")
library(rstanarm)
```

```{r bayesianParamEstimation, echo = FALSE, include = FALSE}
# Fitting a bunch of different glms with different prior intercepts

# Trying model with: Student-T and poisson
# bayesian_model <- stan_glm(
#  NumberStruckActual ~ FlightPhase + Altitude,
#  data = data,
#  family = poisson(link = "log"),
#  prior = student_t(3, 0, 2.5),  # Use Student-t prior instead of normal!
#  prior_intercept = student_t(3, 0, 5),
#  chains = 4, iter = 8000
#)
#summary(bayesian_model)

# Trying model with: Normal and poisson
#bayesian_model_norm <- stan_glm(
#  NumberStruckActual ~ FlightPhase + Altitude,
#  data = data,
#  family = poisson(link = "log"),
#  prior = normal(0, 2.5),
#  prior_intercept = normal(0, 5),
#  chains = 4, iter = 8000 
#)
#summary(bayesian_model_norm)
```

```{r bayesianAltitudeCleanup, echo=FALSE, include=FALSE}
# standardizes altitude 
data$Altitude <- scale(data$Altitude)
#summary(data)

```

```{r}
# Trying model with: Normal and Neg Binomial
# USING THIS ONE
bayesian_model_nb <- stan_glm(
  NumberStruckActual ~ FlightPhase + Altitude,
  data = data,
  family = neg_binomial_2(link = "log"),
  prior = normal(0, 2.5),
  prior_intercept = normal(0, 5),
  chains = 4, iter = 8000, warmup = 500
)
summary(bayesian_model_nb)

```

```{r summary, echo = FALSE, include = TRUE}

# Summarize model output
print(summary(bayesian_model_nb))

# Extract posterior intervals
posterior_intervals <- posterior_interval(bayesian_model_nb, prob = 0.95)
print(posterior_intervals)

# Plot posterior distributions
mcmc_areas(
  as.matrix(bayesian_model_nb),
  pars = c("(Intercept)", 
           "FlightPhaseClimb", 
           "FlightPhaseLanding Roll", 
           "FlightPhaseTake-off run",
           "FlightPhaseDescent",
           "FlightPhaseParked",
           "FlightPhaseTaxi")
) +
  labs(title = "Posterior Distributions of Flight Phase Effects")

# PP CHECK
pp_check(bayesian_model_nb, plotfun = "dens_overlay") +
  labs(title = "Posterior Predictive Check")

```

### Results

Overall, the key findings were as follows. The baseline for `FlightPhase` was the Approach phase. Climb has a posterior mean of near 0 which means no evidence of this phase being associated with more birds struck per incident than the baseline was found. The Landing Roll has a posterior mean of 0.2 which means there is a significant positive association with `NumberStruckActual`. The Take-off run has a posterior mean of 0.2, showing there is also a significant increase in `NumberStruckActual` during this phase. Descent has a posterior mean of -0.3 which means there are predicted to be fewer birds struck per incident relative to the baseline of Approach.

When we analyze the posterior predictive check graph, it seems as though the data is extremely right skewed. This is due to the fact that most bird strike incidents involve very few birds. The prior family of a negative binomial should help with the overdispersion, but it seems like the plot is very skewed. There seems to have a long tail due to extreme values occurring in the dataset. However, the peak at 0 means that the model captures the proportion of flights with zero or very few bird strikes. Even though the graph is not great, it does show that the model is well fit to the data that is present, even though it is skewed.

Based on the prior distributions of landing roll and take off run, these are the flight phases most strongly associated with higher bird strike counts. These findings show that these phases are greater risks for more birds being struck. The confounding variable of `Altitude` should also be taken into account when making data based decisions.

To answer the causal portion of the question, the model will be utilized to assess causal effects. Climb and take-off run have positive coefficients which indicate a higher number of birds struck relative to the baseline. The credible intervals do not cross zero which suggest a strong evidence of a causal relationship. Landing roll is still positive but has a smaller effect than the other two phases. Parked and Taxi have negative coefficients which means there are fewer birds struck compared to baseline, which mean these phases are less likely to contribute to bird strikes causally, since their credible intervals cross zero. From the DAG, we control for the `Altitude`. When we add this into our model, we control for them to isolate the causal impact of the flight phase on number of bird strikes. Since the narrow credible intervals of climb (0.1 - 0.2) and Take-off run (0.14 - 0.22), these are key contributers to the number of birds struck. The Bayesian model not only answered the correlation portion, but also led us to a causal inference conclusion, due to controlling for confounding variables after viewing the DAG.

### Discussion

The findings that were discussed in the results section covered the numerical analysis results, but there are certain limitations to this analysis. Not only can we not actually control for many variables in flights, we have to accept that there is a lot of uncertainty in this analysis. We can quantify our causal analysis and correlations to our numerical analysis with uncertainty factored in, but in reality we can't control for all of those counfounders. The outputs, however, were in line with what I had previously assumed that has the most bird strikes. By confirming it with posterior predictive checks, we can ensure that the priors were fit for the model.
