---
title: "Q4"
author: "Hawkins_Kelsey"
date: "`r Sys.Date()`"
output: pdf_document
---
# ~~~~~ ANALYSIS ~~~~~

# Based on the flight phase ("Climb", "Landing Roll", "Approach", "Take-off run"), which phase is correlated with the largest average number of bird strikes? Is there a casual relationship between flight phase and bird strikes?

# Preliminary EDA and Summary

```{r setup, include=FALSE, echo = FALSE}
# Load libraries
library(tidyverse)
library(ggdag)
library(rstanarm)
library(bayesplot)
library(dagitty)

# Load data
data <- read.csv("Bird_strikes.csv", stringsAsFactors = TRUE)

# Replace NAs with 0
data[is.na(data)] <- 0

```

## Summary Overview
```{r overview, include = TRUE, echo = FALSE}
# Summarize bird strikes by flight phase
bird_strikes_summary <- data %>%
  group_by(FlightPhase) %>%
  summarise(
    MeanStrikes = mean(NumberStruckActual, na.rm = TRUE),
    VarianceStrikes = var(NumberStruckActual, na.rm = TRUE),
    Count = n()
  )

# Print the summary
print(bird_strikes_summary)

# Bar plot of mean bird strikes by flight phase
ggplot(bird_strikes_summary, aes(x = FlightPhase, y = MeanStrikes, fill = FlightPhase)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Average Bird Strikes by Flight Phase",
    x = "Flight Phase",
    y = "Average Number of Bird Strikes"
  ) +
  theme_minimal()

```

## DAG
```{r DAG, include = TRUE, echo = FALSE}
# Construct DAG
dag <- dagitty("
dag {
  AircraftType -> NumberStruckActual
  WeatherConditions -> NumberStruckActual
  TimeOfDay -> NumberStruckActual
  FlightPhase -> NumberStruckActual
}
")

# Plot the DAG
ggdag(dag) +
  theme_minimal() +
  labs(title = "DAG: Confounding Variables for Bird Strikes")

```

# Bayesian Parameter Estimation + Causal Inference
```{r bayesianParamEstimation, echo = FALSE, include = TRUE}
# Fit Bayesian Poisson regression
bayesian_model <- stan_glm(
  NumberStruckActual ~ FlightPhase,
  data = data,
  family = poisson(link = "log"),
  prior = normal(0, 2.5),
  prior_intercept = normal(0, 5),
  chains = 4, iter = 2000
)

# Summarize model output
print(summary(bayesian_model))

# Extract posterior intervals
posterior_intervals <- posterior_interval(bayesian_model, prob = 0.95)
print(posterior_intervals)

# Plot posterior distributions
mcmc_areas(
  as.matrix(bayesian_model),
  pars = c("(Intercept)", 
           "FlightPhaseClimb", 
           "FlightPhaseLanding Roll", 
           "FlightPhaseTake-off run")
) +
  labs(title = "Posterior Distributions of Flight Phase Effects")

# Posterior predictive checks
pp_check(bayesian_model, plotfun = "dens_overlay") +
  labs(title = "Posterior Predictive Check")

```
# ~~~~~ RESULTS ~~~~~

Analysis of Results

1. Posterior Distributions:
The plot of posterior distributions shows the estimated effects of different flight phases on bird strikes.
The intercept represents the baseline effect (likely the log of mean strikes when no specific phase is present).
Positive effects (e.g., for FlightPhaseClimb and FlightPhaseTake-off run) indicate these phases are associated with a higher number of bird strikes relative to the baseline.
Negative effects (e.g., FlightPhaseDescent and FlightPhaseParked) suggest a reduced association compared to the baseline.

2. Credible Intervals:
The 95% credible intervals for FlightPhaseClimb and FlightPhaseTake-off run are entirely positive, indicating significant positive associations with bird strikes.
FlightPhaseParked has a wide credible interval, crossing negative values, which suggests uncertainty and potentially non-significant results.
FlightPhaseTaxi spans both positive and negative values, indicating ambiguity in its effect.

3. Posterior Predictive Check:
The predictive check shows the alignment of simulated data (y_rep) with the observed data (y).
A strong overlap between the two indicates that the model fits the observed data well.
The plot shows extreme values (e.g., very high bird strikes), suggesting potential overdispersion or outliers in the dataset. Consider further diagnostics to assess model fit for extreme counts.

4. MCMC Diagnostics:
Rhat values are all 1.0, indicating good convergence of chains.
Effective sample sizes (n_eff) are high across parameters, ensuring reliable posterior estimates.
Monte Carlo Standard Errors (mcse) are low, indicating precise estimates of posterior means.

Key Takeaways:
Climb and Take-off run phases are strongly associated with higher bird strikes.
The causal impact of flight phases is evident, but further modeling (e.g., including confounders like weather conditions or wildlife size) can refine the results.
The results suggest that bird strike mitigation strategies should focus more on the climb and take-off run phases.

# ~~~~~ DISCUSSION ~~~~~
